%\setcounter{page}{5}

\paragraph{}

%Ao apresentar esta dissertação quero agradecer a todos aqueles que, de alguma forma, contribuíram para a sua concretização particularmente:
%\paragraph{}
%- Ao meu orientador, o Professor Doutor José Orlando Pereira pela orientação, acompanhamento e incentivo que me permitiram levar a cabo esta dissertação;
%\paragraph{}
%- Aos meus colegas do Laboratório de Sistemas Distribuídos pela discussão de ideias, entreajuda e apoio "moral";
%\paragraph{}
%- Aos meus pais pelo apoio e suporte fundamental ao longo de todo o meu percurso;

Firstly, I want to thank Prof. Dr. Rui Oliveira for accepting to be my advisor  and for always pushing me to work harder. His support and guidance was of most value to this dissertation.

Secondly, I would like to thank my family for the constant encouragement throughout my studies.

I also thank all the members of the Distributed Systems Group at University of Minho, for the good working environment provided and for always being available whenever I needed help. A special thank to Ricardo Vilaça, for the constant help, and the patience to listen to me all those days. I big thanks to Pedro Gomes, Nelson Gonçalves, Miguel Borges and Francisco Cruz, for the healthy discussions and brainstorms. 

Thanks to all my friends, for their friendship and for their endless support, especially Miguel Regedor, Roberto Machado, Hugo Marinho, André Santos and Pedro Pereira, who helped me grow as an engineer, a student and a person. 

Also thanks to everyone that read this thesis and contributed with corrections and critics.

Although not personally acquainted I would like to thank Jonathan Ellis for the prompt response both by email and on JIRA.  

Last but not the least I thank Carolina Almeida, who's moral support was vital through the duration of this work. 
 